---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all(quiet = TRUE)
```


# xchan: Cross-Section Based Channels

<!-- badges: start -->
<!-- badges: end -->

The purpose of xchan is to create, widen (erode), and calculate attributes of watercourse geometries. It does this by encoding a channel by its cross sections, which can be planimetric (line segments arranged on a map) or profile (slices into the landscape).

The package concept is inspired by the [**sf**]() package, because xchan encodes both cross section objects (`xsection`) and whole channels (`xchan`), which are simply a list of cross sections. Most functions in xchan start with a common prefix, `xt`, in parallel with the sf package's function prefix, `st`. The `x` stands for "cross" or "cross section".

## Statement of Need

(automation; randomization)

## Target Audience



## Installation

You can install the development version of xchan from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("stochaGBEM/xchan")
```

## Example

The package includes Fraser River demo data that can be used to build a
typical cross-section workflow. Start by loading the package and
unwrapping the packaged DEM:

```{r}
library(xchan)

# Unwrap the packaged DEM to a `terra::SpatRaster` object for use in the package.
dem <- terra::unwrap(demo_dem)
```


Generate planimetric cross sections from the bankline polygon and plot the
stored sampling axis (`xt_generate_plan()` saves it on the channel; use `xt_axis()` to read it):

```{r}
planimetric_cross_sections <- xt_generate_plan(demo_bankline, spacing = 200)
axis_line <- xt_axis(planimetric_cross_sections)

plot(demo_bankline, col = "grey90", border = "grey50")
plot(planimetric_cross_sections, add = TRUE, col = "dodgerblue3")
plot(axis_line, add = TRUE, col = "cadetblue1", lwd = 2)
```


Sample the DEM to create profile cross sections:

```{r}
profile_cross_sections <- xt_generate_profile(
  planimetric_cross_sections,
  dem,
  extent_distance = 200,
  sample_n = 151
)

plot(demo_bankline, col = "grey90", border = "grey50")
plot(profile_cross_sections, add = TRUE, col = "coral")
```


Profile cross sections can also be generated as distance-elevation
objects. Each profile cross section contains a distance-elevation geometry that
can be inspected directly:

```{r}
# Profile cross sections (distance-elevation):
channel <- xt_generate_profile(
  channel = planimetric_cross_sections,
  dem = dem,
  extent_distance = 200,
  sample_n = 151
)

# Plot an example profile cross section
plot(xt_profile_at(profile_cross_sections, 1))
```


Apply a widening scenario to the generated cross sections. In this
example, the channel is widened by 20 metres on the right bank:

```{r}
widened_cross_sections <- xt_widen(
  planimetric_cross_sections,
  dw = 200,
  side = "right"
)

plot(demo_bankline)
plot(widened_cross_sections, add = TRUE)
```


## xchan in the Context of Other R Packages

The following packages are relevant for channel geometries in R.

- The [**centerline**]() package is used to generate a centerline from a bankline polygon.
- The [**sf**]() package is used to manipulate vector-based spatial objects.
- The [**terra**]() package is used to manipulate both vector and raster-based spatial objects.