---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sxchan: simple channel cross sections

<!-- badges: start -->
[![R-CMD-check](https://github.com/stochaGBEM/sxchan/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stochaGBEM/sxchan/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/stochaGBEM/sxchan/branch/main/graph/badge.svg)](https://app.codecov.io/gh/stochaGBEM/sxchan?branch=main)
<!-- badges: end -->

The purpose of sxchan is to create and manipulate watercourse geometries, with a focus on cross sections. Because this package sits on top of the sf package, geometries can also be manipulated in the same way as in the sf package.

The name of the package is inspired by the sf package. Functions in sxchan start with a common prefix, `xt`, which stands for "cross-section type"; this is intended to parallel the sf package's function prefix, `st`, which stands for "spatial type".

## Installation

You can install the development version of sxchan from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("stochaGBEM/sxchan")
```

## Example

```{r}
library(sxchan)
library(sf)
```

### Constructing cross section objects

Flow direction is from the perspective of someone looking downstream, as is the typical convention in hydrology.


Cross section objects can be made using the function `xt_sxc()`, where "sxc" stands for "spatial cross-section column".

1. Simply specify cross section width if you're not interested in the spatial arrangement of the cross sections:

```{r}
cross1 <- xt_sxc(c(8, 7, 5, 6, 5, 8))
plot(cross1)
```

2. Or, you can turn a series of line segments into a cross section object. This time, set a coordinate reference system, too:

```{r}
seg <- st_linestring(matrix(c(0, 1, 0, 1), ncol = 2))
cross2 <- xt_sxc(seg, crs = 3005)
plot(cross2)
```

Retrieve widths:

```{r}
xt_width(cross1)
xt_width(cross2)
```

Because these cross sections are just `sfc` objects from the sf package, you can add features by making it an `sf` object:

```{r}
# st_sf(cross1, swimmability = c(5, 4, 3, 2, 1, 0), roughness = 0.01)
```

### Generating geometries from bankline polygon

sxchan has a built-in demo bankline polygon:

```{r}
plot(demo_bankline)
```

You can generate cross sections and a centerline:

```{r}
demo_cross <- xt_generate_sxc(demo_bankline, n = 50)
plot(demo_bankline)
plot(demo_cross, add = TRUE, col = "blue")
```

You can widen the cross sections, either by a constant amount or one per cross section:

```{r}
plot(demo_bankline)
xt_widen(demo_cross, width = 2) |> 
  plot(add = TRUE, col = "blue")
set.seed(1)
plot(demo_bankline)
xt_widen(demo_cross, width = runif(50, 1, 3)) |> 
  plot(add = TRUE, col = "blue")
```

# sxchan in context

Similar R packages for encoding river channels for the purpose of erosion include the following.

- `rivr` exists to calculate channel hydraulic properties from simple channel geometries.
- `fluvgeo` exists to infer stream properties based on high-resolution terrain data (like a Digital Elevation Model), like dimensions and floodplain inference. It is a computational engine for the FluvialGeomorph ArcGIS toolbox, rather than a standalone tool for creating or manipulating river geometries.

`sxchan` is unique because it is designed to create and manipulate river cross-sections with a focus on channel widening due to bank erosion. This makes it particularly well-suited for conceptual modeling and erosion simulations.

