---
title: "Getting Started with Channels"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Channels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Channel objects are encoded as a list of cross sections. This vignette covers the simple case where cross sections are only planimetric -- that is, they are line segments arranged on a map. Profile geometries are also allowed; see the [_Channels with Profiles_]() vignette to see how.

```{r setup}
library(xchan)
```


## Creating Channels by Hand

The simplest way to make a channel is by specifying a vector of widths. This is useful when the spatial layout of the channel is not important, for example.

```{r}
widths <- c(10, 12, 8, 15, 11, 9)
channel1 <- xt_as_channel(widths)
plot(channel1)
```

There are two main objects in this package:

- A cross section (object of class `"xsection"`)
- A channel (object of class `"xchan"`)

A _cross section_ is a linear slice across a river channel, and in this
package, is spatially encoded through a _planimetric_ ("plan") view and,
optionally, a _profile_ view.

A _channel_ is a list of cross sections, optionally strung together along
a channel _axis_ like a centerline.

```{r}
channel1
```

You could also string those cross sections along a channel axis for a bit more spatial realism. This axis should be encoded as an `sfc` or `sfg` object from the [**sf**]() package.

```{r}
axis <- sf::st_linestring(
  matrix(c(1:10 * 5, 5 * sin(1:10)), ncol = 2),
  dim = "XY"
)
axis
```

String the original widths along the axis; this time, we'll also view the flow direction.

```{r}
channel2 <- xt_as_channel(widths, axis = axis)
plot(channel2, axis = "arrows")
```

For more control, you could specify your own cross sections as bespoke line segments. As usual, make sure these are encoded up as objects with the [**sf**]() package, this time as `sfc` objects.

```{r sfc-input}
library(sf)
lines <- list(
  st_linestring(matrix(c(-2, -1, 5, 0), ncol = 2)),
  st_linestring(matrix(c(1, 0, 5, 0), ncol = 2)),
  st_linestring(matrix(c(3, 2, 3, -1), ncol = 2)),
  st_linestring(matrix(c(5, 4, -1, -2), ncol = 2))
)
lines_sfc <- st_sfc(lines)
lines_sfc
```

```{r}
channel3 <- xt_as_channel(lines_sfc)
plot(channel3)
```

Note that an axis is automatically generated in the direction based on the order the cross sections appear in; if this results in a strange axis, or your line segments are not in order, it's a good idea to take more control of the channel axis -- see the [Channel Axes](#channel-axes) section.

## Creating Channels from Spatial Data

```{r}
plot(demo_bankline)
```

Cross sections can be auto-generated from the bankline polygon. You can specify the cross section spacing directly (`spacing`), or by the desired number of cross sections (`n`). Unlike `xt_as_channel()`, which converts cross-section-like objects into cross sections, `xt_generate_plan()` uses an unrelated object -- the bankline polygon -- to generate cross sections.

```{r}
squamish <- xt_generate_plan(demo_bankline, spacing = 100)
plot(squamish)
```

The algorithm for creating cross sections works by sampling points along a channel axis, and connecting the two banks according to the shortest distance. If no axis is supplied, a centerline is delineated using the [**centerline**]() package. If islands are present, they are ignored when calculating bank-to-bank span; however, the island banks are encoded in the cross sections so that wetted width can be determined (see the Channel Computations vignette for more).

It's worth noting that the algorithm method may produce crossing cross sections. If that's a problem, consider hand picking the cross section locations along the channel axis, or remove the problematic cross sections. For example, remove the first two cross sections, which appear to be artifacts of the sampling procedure.

```{r}
squamish[1:2] <- NULL
```

## Channel Axes

All channels must have an _axis_. This provides two main structural components to the channel.

- It controls flow direction, so that a left and right bank can be identified.
- It determines cross section ordering.

In addition, many channel computations rely heavily on the channel axis -- see the _Channel Computations_ vignette for more.

Running a new axis along the cross sections updates the direction of flow. This is demonstrated below with a new axis directing `channel2`.

```{r}
new_axis <- st_linestring(matrix(c(58, 2, 0, 5), ncol = 2))
xt_axis(channel2) <- new_axis
```

This channel now flow towards the left, the opposite of its previous flow.

```{r}
plot(channel2, axis = "arrows")
```

You can also reverse the flow direction directly:

```{r}
channel2 <- xt_reverse_flow(channel2)
plot(channel2, axis = "arrows")
```

It's important to note that, while a channel is just a list of cross sections, the order of the cross sections in the list does not influence flow order! Try it yourself: does `channel2[c(2, 3, 1)]` change the flow direction and order? The importance of this decoupling is discussed later when cross sections are placed in a data frame.

## Channel Data Frames

Of course, a river is more than its geometry. Any number of characteristics could be relevant for an analysis, such as bed roughness, grain size, or even segment ID. This is why channels are lists, so that you can pair characteristics to each cross section -- ideally, in a data frame.

Consider `channel2`, which has six cross sections. A data frame (tibble) can be constructed to pair cross sections with their characteristics.

```{r}
tibble::tibble(
  cross_section = channel2,
  roughness = c(0.035, 0.035, 0.03, 0.03, 0.03, 0.025),
  gradient = c(0.01, 0.01, 0.02, 0.02, 0.01, 0.01)
)
```

While early development of xchan defined channels as tabular, the mature version only deals with the single column of cross sections because its goals are purely spatial. In addition, it's not obvious how to support the various ways in which operations can be conducted on a table, like expanding joins or split-apply-combine strategies often used in dplyr-like workflows.

## Channel Bankline

Like the [channel axis](#channel-axes), a channel bankline is also stored with a channel object, as an attribute. However, this is attribute is optional, and is only included for convenience. For example, when present, they can be included in a plot. 

You can get and set a bankline polygon with `xt_bankline()`. Here is the polygon that was used to create the spatially-derived channel in the [previous example](#creating-channels-from-spatial-data).

```{r}
xt_bankline(squamish)
```

You can also specify the bankline polygon directly whenever coercing with `xt_as_channel()`.

## Channel Widening

The `xt_widen()` function allows you to widen (erode) a channel. Here is the Squamish River widened by 200 units (meters, in this case).

```{r}
squamish2 <- xt_widen(squamish, dw = 200)
plot(squamish2)
```

By default, erosion occurs evenly on both sides. While you can erode the left bank with `side = "left"` and the right bank with `side = "right"`, you can also control the distribution between left and right banks by a `side_` specifier: `side_both()`, `side_left()`, and `side_right()`. All three functions do the same thing (specify allocation between left and right banks), but provide a different approach to this specification. 

For example, here is a 70-30 left-right split of erosion.

```{r simple-sides}
squamish3 <- xt_widen(squamish, dw = 200, side = side_left(0.7))
xt_width(squamish3)
```

Currently only a single proportion is allowed; a future version of xchan aspires to allow varying proportions along the channel length.

## Exporting Channels

You can export the channel as an object understood by the [**sf**]() package; the channel will be a series of line segments. If there are islands, you have the option of excluding them, resulting in multi-line segments spanning the wet part of the channel.

```{r}
squamish_sf <- xt_as_sfc(squamish, extent = "wetted")
plot(squamish_sf)
```