---
title: "Channel Calculations and Analysis"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Calculations and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```

# Channel Calculations and Analysis

This vignette covers detailed channel calculations including elevation specifications, gradients, heights, and dredging operations.

## Elevation Specifications

`xchan` provides flexible elevation specifications that allow you to reference different parts of the channel.

### Built-in Elevation Specifications

```{r elevation-specs}
# Thalweg elevation (lowest point in channel)
thalweg_elev <- elevation_thalweg()

# Bank elevation (using minimum bank height by default)
bank_elev <- elevation_bank()

# Constant elevation
constant_elev <- elevation_constant(100)

# Column-based elevation (if channel has elevation columns)
# column_elev <- elevation_column("water_surface")

# Bottom elevation (average of channel bottom)
bottom_elev <- elevation_bottom()
```

### Using Elevation Specifications

```{r elevation-usage}
# Get elevation values using specifications
# Note: This requires profile cross-sections
# thalweg_heights <- xt_elevation(channel, reference = thalweg_elev)
# bank_heights <- xt_elevation(channel, reference = bank_elev)
```

## Height Calculations

### Bank Heights

```{r bank-heights}
# Get left and right bank heights
# left_bank_heights <- xt_height_lb(channel)
# right_bank_heights <- xt_height_rb(channel)

# Get aggregated bank heights
# min_heights <- xt_height(channel, .f = min)
# max_heights <- xt_height(channel, .f = max)
# mean_heights <- xt_height(channel, .f = mean)
```

### Channel Height

```{r channel-height}
# Calculate channel height (difference between upper and lower elevations)
# channel_height <- xt_height(channel, lower = elevation_thalweg(), upper = elevation_bank())
```

## Gradient Calculations

### Basic Gradient

```{r basic-gradient}
# Calculate channel gradient
gradient <- xt_gradient(channel)
gradient

# With custom parameters
gradient_custom <- xt_gradient(channel, before = 2, after = 2, complete = FALSE)
gradient_custom
```

### Gradient with Different Elevation References

```{r gradient-elevation}
# Gradient using bank elevation
# bank_gradient <- xt_gradient(channel, elevation = elevation_bank())

# Gradient using thalweg elevation
# thalweg_gradient <- xt_gradient(channel, elevation = elevation_thalweg())
```

## Distance Calculations

### Downstream Distance

```{r downstream-distance}
# Calculate downstream distances
distances <- xt_distance_downstream(channel)
distances

# With a custom axis (sf LINESTRING)
# axis_line <- ...  # user-defined line along the channel
# distances_custom <- xt_distance_downstream(channel, axis = axis_line)
```

## Width Analysis

### Specific Bank Widths

```{r bank-widths}
# Get left and right bank widths
left_widths <- xt_width_lb(channel)
right_widths <- xt_width_rb(channel)

# Compare with total widths
total_widths <- xt_width(channel)
data.frame(
  total = total_widths,
  left = left_widths,
  right = right_widths,
  check = left_widths + right_widths
)
```

## Advanced Elevation Analysis

### Multiple Elevation References

```{r multiple-elevations}
# Compare different elevation references
# elevations <- data.frame(
#   thalweg = xt_elevation(channel, reference = elevation_thalweg()),
#   bank_min = xt_elevation(channel, reference = elevation_bank(min)),
#   bank_max = xt_elevation(channel, reference = elevation_bank(max)),
#   bank_mean = xt_elevation(channel, reference = elevation_bank(mean))
# )
# elevations
```

### Elevation Differences

```{r elevation-differences}
# Calculate various height differences
# bank_to_thalweg <- xt_height(channel, lower = elevation_thalweg(), upper = elevation_bank())
# water_depth <- xt_height(channel, lower = elevation_thalweg(), upper = elevation_constant(water_level))
```

## Channel Analysis Workflow

### Complete Analysis Example

```{r complete-analysis}
# Perform a complete channel analysis
analysis_results <- data.frame(
  station = seq_along(xt_width(channel)),
  width = xt_width(channel),
  left_width = xt_width_lb(channel),
  right_width = xt_width_rb(channel),
  gradient = xt_gradient(channel),
  distance = xt_distance_downstream(channel)
  # Add more calculations as needed
)

analysis_results
```

### Statistical Summaries

```{r statistical-summaries}
# Summary statistics
summary_stats <- data.frame(
  metric = c("Width", "Left Width", "Right Width", "Gradient"),
  mean = c(mean(xt_width(channel)), 
           mean(xt_width_lb(channel)), 
           mean(xt_width_rb(channel)), 
           mean(xt_gradient(channel), na.rm = TRUE)),
  sd = c(sd(xt_width(channel)), 
         sd(xt_width_lb(channel)), 
         sd(xt_width_rb(channel)), 
         sd(xt_gradient(channel), na.rm = TRUE)),
  min = c(min(xt_width(channel)), 
          min(xt_width_lb(channel)), 
          min(xt_width_rb(channel)), 
          min(xt_gradient(channel), na.rm = TRUE)),
  max = c(max(xt_width(channel)), 
          max(xt_width_lb(channel)), 
          max(xt_width_rb(channel)), 
          max(xt_gradient(channel), na.rm = TRUE))
)

summary_stats
```

## Summary

This vignette covered:

- Elevation specifications for flexible elevation references
- Height calculations for banks and channels
- Gradient calculations with different parameters
- Distance calculations along the channel
- Width analysis for specific banks
- Dredging operations with different methods
- Advanced elevation analysis workflows
- Complete channel analysis examples

The next vignette will show you how to work with complex channel geometries including islands and irregular shapes.
