---
title: "Channel Widening and Erosion"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Widening and Erosion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```

# Channel Widening and Erosion

This vignette covers modifying channels through widening operations and calculating erosion volumes.

## Function Factory Pattern

`xchan` uses a function factory pattern to provide flexible and extensible operations. The factories fall into two categories:

**Specification Factories** create functions that define parameters and contexts:
- **Side Factories** (`side_*()`) - specify erosion allocation patterns
- **Elevation Factories** (`elevation_*()`) - specify elevation sources and references

**Operator Factories** create functions that perform actual operations:
- **Dredger Factories** (`dredger_*()`) - perform channel modification operations  
- **Tracer Factories** (`tracer_*()`) - perform bankline tracing operations
- **Extender Factories** (`extender_*()`) - perform topographic frame extensions

This pattern allows for sophisticated parameter capture and composition, such as embedding elevation specifications within dredger functions or extenders.

## Basic Channel Widening

The `xt_widen()` function allows you to modify channel cross-sections by adding width or removing volume.

### Widening by Width

```{r width-widening}
# Create a simple channel
channel <- xt_as_channel(c(10, 12, 8, 15, 11, 9))

# Widen by adding 2 meters to each cross-section
channel_widened <- xt_widen(channel, dw = 2)

# Compare widths
original_widths <- xt_width(channel)
new_widths <- xt_width(channel_widened)
data.frame(original = original_widths, widened = new_widths, difference = new_widths - original_widths)
```

### Widening by Volume

```{r volume-widening}
# Widen by removing 50 cubic meters from each cross-section
# Note: This requires profile cross-sections
# channel_widened_vol <- xt_widen(channel, volume = 50)
```

## Side Specifications

You can control how the widening is distributed between left and right banks.

### Simple Side Specifications

```{r simple-sides}
# Widen both sides equally (default)
channel_both <- xt_widen(channel, dw = 2, side = "both")

# Widen only left side
channel_left <- xt_widen(channel, dw = 2, side = "left")

# Widen only right side  
channel_right <- xt_widen(channel, dw = 2, side = "right")
```

### Using Side Functions (Specification Factories)

Side functions are **specification factories** that create functions defining how erosion should be distributed. They provide more flexible control over erosion distribution:

```{r sides}
# Use built-in side functions
channel_side1 <- xt_widen(channel, dw = 2, side = side_left(0.75))  # 75% left, 25% right
channel_side2 <- xt_widen(channel, dw = 2, side = side_right(0.6))  # 40% left, 60% right
channel_side3 <- xt_widen(channel, dw = 2, side = side_both(0.7, 0.3))  # 70% left, 30% right
```

### Custom Side Functions

You can create custom side functions (specification factories) for specific scenarios:

```{r custom-side}
# Create a custom side function that varies along the channel
custom_side <- function(channel) {
  n <- xt_n_sections(channel)
  # Gradually shift from left to right bank erosion
  seq(0.8, 0.2, length.out = n)
}

channel_custom <- xt_widen(channel, dw = 2, side = custom_side)
```

## Erosion Calculations

### Volume from Width

Calculate how much sediment volume would be removed for a given width increase:

```{r erosion-volume}
# Calculate erosion volume for 2-meter width increase
volumes <- xt_erosion_volume(channel, width = 2, side = "both")
volumes

# With custom side function
volumes_custom <- xt_erosion_volume(channel, width = 2, side = side_left(0.75))
volumes_custom
```

### Width from Volume

Calculate how much width increase would result from removing a given volume:

```{r erosion-width}
# Calculate width increase for 50 cubic meter removal
# Note: This requires profile cross-sections
# widths <- xt_erosion_width(channel, volume = 50, side = "both")
# widths
```

## Advanced Widening Scenarios

### Variable Widening

```{r variable-widening}
# Widen by different amounts at each cross-section
variable_widths <- c(1, 2, 1.5, 3, 2.5, 1)
channel_variable <- xt_widen(channel, dw = variable_widths, side = "both")

# Compare results
data.frame(
  original = xt_width(channel),
  variable = xt_width(channel_variable),
  added = variable_widths
)
```

### Complex Erosion Patterns

```{r complex-erosion}
# Create a side function based on channel characteristics
# For example, more erosion on outside of bends
bend_side <- function(channel) {
  n <- xt_n_sections(channel)
  # Simulate a meandering pattern
  sin_pattern <- sin(seq(0, 2*pi, length.out = n))
  # Convert to proportions (0.3 to 0.7)
  0.5 + 0.2 * sin_pattern
}

channel_bend <- xt_widen(channel, dw = 2, side = bend_side)
```

## Visualizing Widening

```{r visualize-widening}
# Plot original and widened channels
par(mfrow = c(1, 2))
plot(channel, view = "plan", main = "Original Channel")
plot(channel_widened, view = "plan", main = "Widened Channel")
par(mfrow = c(1, 1))
```

## Topographic Frame Extension

Before widening channels, you may need to extend the topographic frame beyond the banks to provide space for erosion calculations. This is especially important for channels created from simple width specifications.

### Using Extenders (Operator Factories)

Extenders are **operator factories** that create functions for extending the topographic frame:

```{r extenders}
# Flat extension on both sides
channel_extended <- xt_extend_frame(channel, extender = extender_flat(extent = 20))

# Sloped extension on both sides
channel_sloped <- xt_extend_frame(channel, extender = extender_slope(extent = 30, slope = 0.02))

# Different extensions on each side
channel_asymmetric <- xt_extend_frame_left(channel, extender = extender_slope(extent = 30, slope = 0.05))
channel_asymmetric <- xt_extend_frame_right(channel_asymmetric, extender = extender_flat(extent = 20))
```

### Using Different Elevation References

You can specify different elevation references for extensions:

```{r elevation-references}
# Extend from bank elevation
channel_bank <- xt_extend_frame(channel, extender = extender_flat(extent = 20, elevation = elevation_bank()))

# Extend from leftmost point elevation
channel_left <- xt_extend_frame(channel, extender = extender_flat(extent = 20, elevation = elevation_topo_left()))

# Extend from rightmost point elevation
channel_right <- xt_extend_frame(channel, extender = extender_flat(extent = 20, elevation = elevation_topo_right()))
```

## Summary

This vignette covered:

- Basic channel widening by width and volume
- Simple and complex side specifications using side functions
- Custom side function creation for specific scenarios
- Erosion volume and width calculations
- Variable widening patterns
- Topographic frame extension for realistic simulations

The next vignette will show you how to perform detailed channel calculations and analysis.
