---
title: "Channel Objects and Generation"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Objects and Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
library(sf)
```

# Channel Objects and Generation

This vignette covers creating channel objects from different data sources and generating derived geometries.

## Creating Channels from Different Inputs

Use [xt_as_channel()] to **coerce** common inputs (widths, `sfc`, or a data
frame whose rows you pair with planimetric lines via `plan_col` and optional
`profile_col`) into a channel table. Storage is always a single **`xsection`**
list-column (plan plus optional profile per row); [channel_plan()] and
[channel_profile()] are derived views of that column, not separate stored
fields.

When you already have `sfc_LINESTRING` transects (and optionally a same-length
list of `xs_profile` objects), pass them to [xt_as_channel()] --- the `sfc`
method binds plan and profiles into one `xsection` column per row, similar in
spirit to [sf::st_sf()]. Set CRS on the geometries with `sf::st_set_crs()`
before constructing if needed, or pass `crs` to [xt_as_channel()].

When `xt_as_channel()` **creates** the table from a width vector or from `sfc`
geometries, you can pass extra channel columns as further named arguments (like
extra columns in [base::data.frame()]). When you **already** have a data frame,
add any extra variables as columns on that frame first, then call
`xt_as_channel()` with `plan_col` / optional `profile_col`.

When both plan and profile inputs are supplied, construction checks that each
profile spans the same horizontal distance as its planimetric cross section
(width from `xt_width()` matches between the two). The same [xt_width()]
function also accepts a single `xs_profile` object if you need that width in
isolation.

### From Numeric Vectors (Widths)

```{r numeric-input}
# Create channel from width vector
widths <- c(10, 12, 8, 15, 11, 9)
channel <- xt_as_channel(widths)
xt_width(channel)
```

### From Spatial Objects (sfc)

```{r sfc-input}
# Create planimetric cross-sections
library(sf)

# Create simple line geometries
lines <- list(
  st_linestring(matrix(c(0, 0, 10, 0), ncol = 2)),
  st_linestring(matrix(c(0, 0, 12, 0), ncol = 2)),
  st_linestring(matrix(c(0, 0, 8, 0), ncol = 2))
)
plan_sfc <- st_sfc(lines)

# Create channel from sfc
channel <- xt_as_channel(plan_sfc)
xt_width(channel)
```

### From Data Frames

Put metadata and other fields on the data frame **before** coercing; optional
CRS can be set with `xt_as_channel(..., crs = )` or by setting CRS on the
geometries beforehand (the channel stores CRS once on the cross-section column).

```{r df-input}
# Data frame with a column of LINESTRING geometries (coercion reads `plan_col`)
df <- data.frame(
  station = c(0, 100, 200),
  roughness = c(0.03, 0.035, 0.025)
)
df$plan <- plan_sfc

channel <- xt_as_channel(df, plan_col = "plan")
channel
# Planimetric lines: channel_plan(channel)
```

## Generating Channels from Banklines

### From Bankline Polygons

```{r bankline-generation}
# Bankline polygon (package data)
plot(fraser_bankline, main = "Fraser demo bankline")

# Channel table with one xsection per row (plan from bankline sampling)
channel <- xt_generate_plan(fraser_bankline, n = 20)
plot(channel_plan(channel), add = TRUE, col = "blue")
```

### From DEM Data

```{r dem-generation}
# Generate profile cross-sections from DEM
# This would require a DEM raster
# profile <- xt_generate_profile(channel, dem = my_dem)

# For now, we'll create a simple example
# channel_with_profile <- xt_as_channel(df, plan_col = "plan", profile_col = "profile")
```

## Generating Derived Geometries

### Channel axis

```{r channel-axis}
# Trace an axis along the channel from cross-section midpoints
centerline <- xt_trace_centerline(channel)
plot(channel)
plot(centerline, add = TRUE, col = "red", lwd = 2)
```

## Correcting Streamflow Direction

Sometimes the auto-generated streamflow direction may be incorrect:

```{r reverse-flow}
# Check current flow direction
plot(channel)

# Reverse flow direction if needed
channel_reversed <- xt_reverse_flow(channel)
plot(channel_reversed)
```

## Spatial Outputs

### Converting to sf Objects

```{r spatial-outputs}
# Convert channel to sf objects (plan always; profile only if present)
plan_sf <- xt_as_sf(channel, what = "plan")
class(plan_sf)
plot(plan_sf)

# profile_sf <- xt_as_sf(channel, what = "profile")
# three_d_sf <- xt_as_sf(channel, what = "3d")
```

### Converting to sfc Objects

```{r sfc-outputs}
# Convert to sfc objects (geometry only)
plan_sfc <- xt_as_sfc(channel, what = "plan")
# profile_sfc <- xt_as_sfc(channel, what = "profile")

st_length(plan_sfc)
```

## Channel Validation

Check if objects are valid channels:

```{r validation}
# Check if object is a channel (table or xchan column container)
xt_is_channel(channel)

# Check if channel has profile cross sections
xt_has_profile(channel)

# Validate planimetric cross-sections
validate_plan(channel_plan(channel))
```

## Working with Channel Components

### Accessing Channel Columns

```{r column-access}
# Get planimetric cross-sections
plan <- channel_plan(channel)

# Get profile cross-sections (if they exist)
# profile <- channel_profile(channel)

# Replace plan geometries with a validated `sfc_LINESTRING`, for example:
# channel_plan(channel) <- updated_plan_sfc
```

### Plotting One Profile Cross Section

When you want to inspect one profile cross section from a channel, use
`xt_profile_at()` to extract a single `xs_profile`, then call `plot()` on that
profile object:

```{r one-profile-plot}
# Extract one profile (row order) and plot it (requires profile geometry)
coords <- matrix(c(-2, 10, 0, 8, 2, 10), ncol = 2, byrow = TRUE)
xs <- xchan:::new_profile(coords, bankpoints = c(-2, 2))
ch_prof <- xt_as_channel(c(4, 4), profile = list(xs, xs), crs = 3005)
plot(xt_profile_at(ch_prof, 1))
```

## Summary

This vignette covered:

- Creating channels via coercion with `xt_as_channel()`
- Generating channels from banklines and DEMs
- Creating derived geometries (channel axis)
- Correcting streamflow direction
- Converting channels to spatial objects
- Validating and accessing channel components

The next vignette will show you how to modify channels through widening and erosion operations.
