---
title: "Channel Modification"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Widening and Erosion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```

## Channel Widening and Erosion

This vignette covers modifying channels through widening operations and calculating erosion volumes. By default, erosion applies to both banks equally; see the [## Side Specifications](## Side Specifications) section below for changing that.

The `xt_widen()` function allows you to modify channel cross-sections by adding width or removing volume.

```{r width-widening}
# Create a simple channel
channel <- xt_as_channel(c(10, 12, 8, 15, 11, 9))

# Widen by adding 2 meters to each cross-section
channel_widened <- xt_widen(channel, dw = 2)

plot(channel_widened, col = "red")
plot(channel, add = TRUE)
```

This operation is vectorized, so different amounts of erosion can be specified for each cross section.

```{r width-widening}
# Create a simple channel
channel <- xt_as_channel(c(10, 12, 8, 15, 11, 9))

# Widen by adding 2 meters to each cross-section
channel_widened <- xt_widen(channel, dw = c(5, 3, 7, 0, 0, 0))

xt_width(channel_widened)
```

If a profile view is available, you can also widen by volume -- provided there is enough volume to erode in the first place.

Sometimes it's useful to know how much volume of erosion would result from a given change in width, or vice versa, the width change resulting from a volume change. Note that because volume is in squared length units, not cubed, because it corresponds to a cross section.

```{r}
xt_erosion_width(channel, dv = units::set_units(10, "m2"))
xt_erosion_volume(channel, dw = units::set_units(100, "m"))
```


## Side Specifications

You can control how the widening is distributed between left and right banks. Simple side specifications:

```{r simple-sides}
# Widen both sides equally (default)
channel_both <- xt_widen(channel, dw = 2, side = "both")

# Widen only left side
channel_left <- xt_widen(channel, dw = 2, side = "left")

# Widen only right side
channel_right <- xt_widen(channel, dw = 2, side = "right")
```

Under the hood, these arguments call `side_both()`, `side_left()`, and `side_right()` -- functions that provide more control as to how erosion is allocated between the left and right banks. All three functions return the same data structure, but are included for convenience. Currently only a single proportion is allowed; a future version of **xchan** aspires to allow varying proportions along the channel length.

For example, here is a 70-30 left-right split of erosion.

```{r simple-sides}
channel_both <- xt_widen(channel, dw = 2, side = side_left(0.7))
xt_width(channel_both)
```

I could also specify both proportions in `side_both()` as a way of insisting that both proportions sum to 1, but the result is the same.

```{r simple-sides}
channel_both <- xt_widen(
  channel,
  dw = 2,
  side = side_both(prop_left = 0.7, prop_right = 0.3)
)
xt_width(channel_both)
```

