---
title: "Channel Computations"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Computations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

While the [Getting Started with Channels](basic_channels.html) and [Channels with Profiles](channels_with_profiles.html) vignettes cover creating and eroding channels, this vignette covers calculations on channels.

First, make a channel for the Squamish River (same workflow as in the [Channels with Profiles](channels_with_profiles.html) vignette).

```{r setup}
library(xchan)
library(terra)
squamish <- xt_generate_plan(squamish_bankline, spacing = 500)
squamish <- xt_generate_profile(squamish, unwrap(squamish_dem), sample_freq = 10)
plot(squamish)
```

## Distance Downstream

<span class="badge rounded-pill text-bg-secondary">Planimetric cross sections only</span>

Downstream distances can be calculated along the channel axis, where "0" is the canonical distance for the most upstream cross section.

```{r distance}
dist <- xt_distance_downstream(squamish)
head(dist)
```

As with most computational functions, `xt_distance_downstream()` has an optional `axis` argument if you wanted to use a different axis than the stored one.


## Elevations

<span class="badge rounded-pill text-bg-secondary">Profile cross sections</span>

Channel elevation can be defined in a few different ways. Some references use a single point, such as the thalweg elevation.

```{r}
head(xt_elevation(squamish, reference = elevation_thalweg()))
```

Other references summarize multiple points with a function. For example, `elevation_bank()` applies a function to the two outer banks; by default, it returns the lower of the two.

```{r}
head(xt_elevation(squamish, reference = elevation_bank()))
```

Another option is to summarize elevations along the wetted bed with `elevation_bed()`. Here we take the median.

```{r}
head(xt_elevation(squamish, reference = elevation_bed(median)))
```

To learn more about the different elevation references, see the documentation at `?elevations`.

## Gradient

<span class="badge rounded-pill text-bg-secondary">Profile cross sections</span>

A useful way of approximating flow gradient is differencing some aspect of the topography between cross sections along the channel axis. The `xt_gradient()` function calculates this differential using a sliding window approach based on number of cross sections before and after each cross section.

For example, we can calculate the gradient as the slope of the banks three cross sections upstream and downstream of each cross section. Use the [Elevation](#elevations) section to specify the elevation reference.

```{r}
grad <- xt_gradient(
  squamish,
  before = 3,
  after = 3,
  elevation = elevation_thalweg()
)
head(grad)
```

By default, truncated windows are not allowed, which is why the first three cross sections are `NA`. This behaviour can be changed with the `complete` argument.

We can combine this with the `xt_distance_downstream()` function to plot the gradient along the channel axis.

```{r}
plot(dist, grad)
lines(dist, grad, type = "l")
```

While you can make this plot smoother by increasing the window size, it's worth considering combining this with a smoother function like `stats::loess()`.

## Other Computations

If these computations don't cover your needs, you can take a manual approach by first exporting the channel with `xt_as_sfc()` and then using the [**sf**](https://r-spatial.github.io/sf/) package to conduct the calculations. But of course, not until after [submitting an Issue](https://github.com/fluvtools/xchan/issues) on the xchan GitHub repository! Your feedback is greatly appreciated.
