---
title: "Channel Computations"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Computations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Create a custom output hook to truncate vector outputs
knitr::knit_hooks$set(output = function(x, options) {
  # Check if our custom option 'truncate_vector' is enabled (we will set it to TRUE by default)
  if (isTRUE(options$truncate_vector)) {
    # Clean up the raw string output into R lines
    lines <- readLines(textConnection(x))

    # We target standard vector outputs (usually starting with [1])
    if (any(grepl("^\\[\\d+\\]", lines))) {
      # Extract the actual data elements, discarding the R header/bracket notation
      clean_elements <- unlist(strsplit(
        gsub("^\\[\\d+\\]\\s+", "", lines),
        "\\s+"
      ))

      total_elements <- length(clean_elements)

      if (total_elements > 6) {
        remaining <- total_elements - 6
        # Take the first 6, format them nicely, and append the dynamic message
        truncated_data <- paste(clean_elements[1:6], collapse = " ")

        # Re-wrap it in R's classic print style
        x <- paste0(
          "[1] ",
          truncated_data,
          "\n... ",
          remaining,
          " more elements.\n"
        )
      }
    }
  }

  # Pass the (modified or unmodified) output back to knitr's default handler
  knitr::knit_hooks$get("output")(x, options)
})

# Set this behavior as the global default for the whole document
knitr::opts_chunk$set(truncate_vector = TRUE)
```

While the [Getting Started with Channels]() and [Channels with Profiles]()
vignettes cover creating and eroding channels, this vignette covers calculations on channels.

First, make a channel for the Squamish River (same workflow as in the [Channels with Profiles]() vignette).

```{r setup}
library(xchan)
library(terra)
squamish <- xt_generate_plan(demo_bankline, spacing = 100)
squamish <- xt_generate_profile(squamish, unwrap(demo_dem), sample_freq = 2)
plot(squamish)
```

## Distance Downstream

\[Only needs planimetric cross sections.\]

Downstream distances can be calculated along the channel axis, where "0" is the canonical distance for the most upstream cross section.

```{r distance}
dist <- xt_distance_downstream(squamish)
dist
```

As with most computational functions, `xt_distance_downstream()` has an optional `axis` argument if you wanted to use a different axis than the stored one.


## Elevations

\[Needs profile cross sections.\]

Channel elevation can be calculated using a number of different reference concepts.
Some reference concepts select a specific point, like the thalweg elevations.

```{r}
xt_elevation(squamish, reference = elevation_thalweg())
```

Other reference concepts involve combining mulitple points with a function,
like `elevation_bank()`, which applies a function to the two outer banks -- by default, the minimum.

```{r}
xt_elevation(squamish, reference = elevation_bank())
```

Another option is to combine the bed elevations with `elevation_bed()`; here, taking the median elevation.

```{r}
xt_elevation(squamish, reference = elevation_bank(median))
```

To learn more about the different elevation reference concepts, see the documentation at `?elevations`.

## Gradient

\[Needs profile cross sections.\]

A useful way of approximating flow gradient is differencing some aspect of the topography between cross sections along the channel axis.
The `xt_gradient()` function calculates this differential using a sliding window approach based on number of cross sections before and after each cross section.

For example, we can calculate the gradient as the slope of the banks three cross sections upstream and downstream of each cross section. Use the [Elevation](#elevations) concept to specify the elevation reference.

```{r}
grad <- xt_gradient(
  squamish,
  before = 3,
  after = 3,
  elevation = elevation_thalweg()
)
grad
```

By default, truncated windows are not allowed, which is why the first three cross sections are `NA`. This behaviour can be changed with the `complete` argument.

We can combine this with the `xt_distance_downstream()` function to plot the gradient along the channel axis.

```{r}
plot(dist, grad)
lines(dist, grad, type = "l")
```

While you can make this plot smoother by increasing the window size, it's worth considering combining this with a smoother function like `stats::loess()`.

## Other Computations

If these computations don't cover your needs, you can take a manual approach by first exporting the channel with `xt_as_sfc()` and then using the [**sf**]() package to conduct the calculations. But of course, not until after [submitting an Issue]() on the xchan github repository! Your feedback is greatly appreciated.