---
title: "Channel Objects and Generation"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Objects and Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
library(sf)
```

# Channel Objects and Generation

This vignette covers creating channel objects from different data sources and generating derived geometries.

## Creating Channels from Different Inputs

Use [xt_as_channel()] to **coerce** common inputs (widths, `sfc`, or a data
frame with a **plan** column and optional profile column) to a channel.
Use [xt_channel()] when you already have `sfc_LINESTRING` (and optionally
`xs_profile` objects) ready to bind into a table, similar to [sf::st_sf()]. Set CRS
on plan geometries with `sf::st_set_crs()` before constructing if needed.

When `xt_as_channel()` **creates** the table from a width vector or from `sfc`
geometries, you can pass extra channel columns as further named arguments (like
extra columns in `xt_channel()`). When you **already** have a data frame, add any
extra variables as columns on that frame first, then call `xt_as_channel()` ---
or use `xt_channel()` to bind plan, profile, and other columns in one step.

When both plan and profile columns are supplied, construction checks that each
profile spans the same horizontal distance as its planimetric cross section
(width from `xt_width()` matches between the two). The same [xt_width()]
function also accepts a single `xs_profile` object if you need that width in
isolation.

### From Numeric Vectors (Widths)

```{r numeric-input}
# Create channel from width vector
widths <- c(10, 12, 8, 15, 11, 9)
channel <- xt_as_channel(widths)
xt_width(channel)
```

### From Spatial Objects (sfc)

```{r sfc-input}
# Create planimetric cross-sections
library(sf)

# Create simple line geometries
lines <- list(
  st_linestring(matrix(c(0, 0, 10, 0), ncol = 2)),
  st_linestring(matrix(c(0, 0, 12, 0), ncol = 2)),
  st_linestring(matrix(c(0, 0, 8, 0), ncol = 2))
)
plan_sfc <- st_sfc(lines)

# Create channel from sfc
channel <- xt_as_channel(plan_sfc)
xt_width(channel)
```

### From Data Frames

Put metadata and other fields on the data frame **before** coercing; optional
CRS can be set on the plan column with `xt_as_channel(..., crs = )` or by
setting CRS on the geometries beforehand.

```{r df-input}
# Data frame whose `plan` column holds LINESTRING geometries
df <- data.frame(
  station = c(0, 100, 200),
  roughness = c(0.03, 0.035, 0.025)
)
df$plan <- plan_sfc

channel <- xt_as_channel(df, plan_col = "plan")
channel
```

## Generating Channels from Banklines

### From Bankline Polygons

```{r bankline-generation}
# Bankline polygon (package data)
bl <- st_sfc(demo_bankline, crs = 3005)
plot(bl, main = "Demo Bankline")

# Channel whose plan column holds cross-section LINESTRINGs
channel <- xt_generate_plan(bl, n = 20)
plot(xt_column_plan(channel), add = TRUE, col = "blue")
```

### From DEM Data

```{r dem-generation}
# Generate profile cross-sections from DEM
# This would require a DEM raster
# profile <- xt_generate_profile(channel, dem = my_dem)

# For now, we'll create a simple example
# channel_with_profile <- xt_channel(channel, profile_col = profile)
```

## Generating Derived Geometries

### Channel axis

```{r channel-axis}
# Trace an axis along the channel from cross-section midpoints
centerline <- xt_trace_centerline(channel)
plot(channel, view = "plan")
plot(centerline, add = TRUE, col = "red", lwd = 2)
```

### Banklines

```{r banklines, eval=FALSE}
# Bankline tracers are currently excluded from the package build.
# banklines <- xt_trace_banks(channel)
# plot(channel, view = "plan")
# plot(banklines, add = TRUE, col = "green", lwd = 2)
```

## Correcting Streamflow Direction

Sometimes the auto-generated streamflow direction may be incorrect:

```{r reverse-flow}
# Check current flow direction
plot(channel, view = "plan")

# Reverse flow direction if needed
channel_reversed <- xt_reverse_flow(channel)
plot(channel_reversed, view = "plan")
```

## Spatial Outputs

### Converting to sf Objects

```{r spatial-outputs}
# Convert channel to sf objects
plan_sf <- xt_as_sf(channel, what = "plan")
profile_sf <- xt_as_sf(channel, what = "profile")
# three_d_sf <- xt_as_sf(channel, what = "3d")  # if both plan and profile exist

# Work with sf objects
class(plan_sf)
plot(plan_sf)
```

### Converting to sfc Objects

```{r sfc-outputs}
# Convert to sfc objects (geometry only)
plan_sfc <- xt_as_sfc(channel, what = "plan")
profile_sfc <- xt_as_sfc(channel, what = "profile")

# Use in spatial operations
st_length(plan_sfc)
```

## Channel Validation

Check if objects are valid channels:

```{r validation}
# Check if object is a channel
is_channel(channel)

# Check if channel has specific components
xt_has_plan(channel)
xt_has_profile(channel)

# Validate planimetric cross-sections
xt_validate_plan(xt_column_plan(channel))
```

## Working with Channel Components

### Accessing Channel Columns

```{r column-access}
# Get planimetric cross-sections
plan <- xt_column_plan(channel)

# Get profile cross-sections (if they exist)
# profile <- xt_column_profile(channel)

# Replace plan geometries with a validated `sfc_LINESTRING`, for example:
# xt_column_plan(channel) <- updated_plan_sfc
```

## Summary

This vignette covered:

- Creating channels via coercion (`xt_as_channel()`) or explicit assembly (`xt_channel()`)
- Generating channels from banklines and DEMs
- Creating derived geometries (channel axis, banklines)
- Correcting streamflow direction
- Converting channels to spatial objects
- Validating and accessing channel components

The next vignette will show you how to modify channels through widening and erosion operations.
