---
title: "Channel Objects and Generation"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Objects and Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sxchan)
library(sf)
```

# Channel Objects and Generation

This vignette covers creating channel objects from different data sources and generating derived geometries.

## Creating Channels from Different Inputs

The `xt_channel()` function is an S3 generic that can work with various input types.

### From Numeric Vectors (Widths)

```{r numeric-input}
# Create channel from width vector
widths <- c(10, 12, 8, 15, 11, 9)
channel <- xt_channel(widths)
xt_width(channel)
```

### From Spatial Objects (sfc)

```{r sfc-input}
# Create planimetric cross-sections
library(sf)

# Create simple line geometries
lines <- list(
  st_linestring(matrix(c(0, 0, 10, 0), ncol = 2)),
  st_linestring(matrix(c(0, 0, 12, 0), ncol = 2)),
  st_linestring(matrix(c(0, 0, 8, 0), ncol = 2))
)
plan_sfc <- st_sfc(lines)

# Create channel from sfc
channel <- xt_channel(plan_sfc)
xt_width(channel)
```

### From Data Frames

```{r df-input}
# Create data frame with plan and profile columns
df <- data.frame(
  station = c(0, 100, 200, 300, 400, 500),
  width = c(10, 12, 8, 15, 11, 9),
  roughness = c(0.03, 0.035, 0.025, 0.04, 0.03, 0.025)
)

# Create channel specifying column names
channel <- xt_channel(df, plan_col = "width")
channel
```

## Generating Channels from Banklines

### From Bankline Polygons

```{r bankline-generation}
# Use demo bankline
plot(demo_bankline, main = "Demo Bankline")

# Generate planimetric cross-sections from bankline
plan <- xt_generate_plan(demo_bankline, n = 20)
plot(plan, add = TRUE, col = "blue")

# Create channel from generated plan
channel <- xt_channel(plan)
```

### From DEM Data

```{r dem-generation}
# Generate profile cross-sections from DEM
# This would require a DEM raster
# profile <- xt_generate_profile(channel, dem = my_dem)

# For now, we'll create a simple example
# channel_with_profile <- xt_channel(channel, profile_col = profile)
```

## Generating Derived Geometries

### Centerlines

```{r centerline}
# Generate centerline from channel
centerline <- xt_trace_centerline(channel)
plot(channel, view = "plan")
plot(centerline, add = TRUE, col = "red", lwd = 2)
```

### Banklines

```{r banklines}
# Generate banklines from channel
banklines <- xt_trace_banks(channel)
plot(channel, view = "plan")
plot(banklines, add = TRUE, col = "green", lwd = 2)
```

## Correcting Streamflow Direction

Sometimes the auto-generated streamflow direction may be incorrect:

```{r reverse-flow}
# Check current flow direction
plot(channel, view = "plan")

# Reverse flow direction if needed
channel_reversed <- xt_reverse_flow(channel)
plot(channel_reversed, view = "plan")
```

## Spatial Outputs

### Converting to sf Objects

```{r spatial-outputs}
# Convert channel to sf objects
plan_sf <- xt_as_sf(channel, type = "plan")
profile_sf <- xt_as_sf(channel, type = "profile")
# 3d_sf <- xt_as_sf(channel, type = "3d")  # if both plan and profile exist

# Work with sf objects
class(plan_sf)
plot(plan_sf)
```

### Converting to sfc Objects

```{r sfc-outputs}
# Convert to sfc objects (geometry only)
plan_sfc <- xt_as_sfc(channel, type = "plan")
profile_sfc <- xt_as_sfc(channel, type = "profile")

# Use in spatial operations
st_length(plan_sfc)
```

## Channel Validation

Check if objects are valid channels:

```{r validation}
# Check if object is a channel
is_channel(channel)

# Check if channel has specific components
xt_has_plan(channel)
xt_has_profile(channel)

# Validate planimetric cross-sections
xt_validate_plan(plan)
```

## Working with Channel Components

### Accessing Channel Columns

```{r column-access}
# Get planimetric cross-sections
plan <- xt_column_plan(channel)

# Get profile cross-sections (if they exist)
# profile <- xt_column_profile(channel)

# Set new planimetric cross-sections
new_plan <- plan * 1.2  # Scale by 20%
xt_column_plan(channel) <- new_plan
```

## Summary

This vignette covered:

- Creating channels from different input types (numeric, sfc, data.frame)
- Generating channels from banklines and DEMs
- Creating derived geometries (centerlines, banklines)
- Correcting streamflow direction
- Converting channels to spatial objects
- Validating and accessing channel components

The next vignette will show you how to modify channels through widening and erosion operations.
