---
title: "Channel Widening and Erosion"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channel Widening and Erosion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```

# Channel Widening and Erosion

This vignette covers modifying channels through widening operations and calculating erosion volumes.

## Function Factory Pattern

`xchan` uses a function factory pattern to provide flexible and extensible operations. The factories fall into two categories:

**Specification Factories** create functions that define parameters and contexts:
- **Side Factories** (`side_*()`) - specify erosion allocation patterns
- **Elevation Factories** (`elevation_*()`) - specify elevation sources and references

Core operations such as `xt_widen()` combine these specifications with channel geometry.

This pattern allows for sophisticated parameter capture and composition, such as embedding elevation references in widening or erosion calculations.

## Basic Channel Widening

The `xt_widen()` function allows you to modify channel cross-sections by adding width or removing volume.

### Widening by Width

```{r width-widening}
# Create a simple channel
channel <- xt_as_channel(c(10, 12, 8, 15, 11, 9))

# Widen by adding 2 meters to each cross-section
channel_widened <- xt_widen(channel, dw = 2)

# Compare widths
original_widths <- xt_width(channel)
new_widths <- xt_width(channel_widened)
data.frame(original = original_widths, widened = new_widths, difference = new_widths - original_widths)
```

### Widening by Volume

```{r volume-widening}
# Widen by removing 50 cubic meters from each cross-section
# Note: This requires profile cross-sections
# channel_widened_vol <- xt_widen(channel, volume = 50)
```

## Side Specifications

You can control how the widening is distributed between left and right banks.

### Simple Side Specifications

```{r simple-sides}
# Widen both sides equally (default)
channel_both <- xt_widen(channel, dw = 2, side = "both")

# Widen only left side
channel_left <- xt_widen(channel, dw = 2, side = "left")

# Widen only right side  
channel_right <- xt_widen(channel, dw = 2, side = "right")
```

### Using Side Functions (Specification Factories)

Side functions are **specification factories** that create functions defining how erosion should be distributed. They provide more flexible control over erosion distribution:

```{r sides}
# Use built-in side functions
channel_side1 <- xt_widen(channel, dw = 2, side = side_left(0.75))  # 75% left, 25% right
channel_side2 <- xt_widen(channel, dw = 2, side = side_right(0.6))  # 40% left, 60% right
channel_side3 <- xt_widen(channel, dw = 2, side = side_both(0.7, 0.3))  # 70% left, 30% right
```

### Varying proportions along the channel

`side_left()`, `side_right()`, and `side_both()` recycle scalar proportions to
every cross section. For section-wise control, build an `xchan_side` object
whose `left` (and `right`) entries are numeric vectors of length
`xt_n_sections(channel)` that sum to 1 at each section:

```{r custom-side}
n <- xt_n_sections(channel)
prop_left <- seq(0.8, 0.2, length.out = n)
custom_side <- structure(
  list(left = prop_left, right = 1 - prop_left),
  class = "xchan_side"
)

channel_custom <- xt_widen(channel, dw = 2, side = custom_side)
```

## Erosion Calculations

### Volume from Width

`xt_erosion_volume()` estimates volume removed for a given width increase; it
requires profile cross sections and a widening that stays within each profile’s
vertical extent (see the function documentation). With a channel that has
profiles, for example:

```{r erosion-volume, eval=FALSE}
# volumes <- xt_erosion_volume(channel_with_profiles, dw = 1, side = "both")
# volumes_custom <- xt_erosion_volume(
#   channel_with_profiles,
#   dw = 1,
#   side = side_left(0.75)
# )
```

### Width from Volume

Calculate how much width increase would result from removing a given volume:

```{r erosion-width}
# Calculate width increase for 50 cubic meter removal
# Note: This requires profile cross-sections
# widths <- xt_erosion_width(channel, volume = 50, side = "both")
# widths
```

## Advanced Widening Scenarios

### Variable Widening

```{r variable-widening}
# Widen by different amounts at each cross-section
variable_widths <- c(1, 2, 1.5, 3, 2.5, 1)
channel_variable <- xt_widen(channel, dw = variable_widths, side = "both")

# Compare results
data.frame(
  original = xt_width(channel),
  variable = xt_width(channel_variable),
  added = variable_widths
)
```

### Complex Erosion Patterns

```{r complex-erosion}
# Section-wise proportions (e.g. mimicking more erosion on one side along a bend)
n <- xt_n_sections(channel)
sin_pattern <- sin(seq(0, 2 * pi, length.out = n))
prop_left <- 0.5 + 0.2 * sin_pattern
bend_side <- structure(
  list(left = prop_left, right = 1 - prop_left),
  class = "xchan_side"
)

channel_bend <- xt_widen(channel, dw = 2, side = bend_side)
```

## Visualizing Widening

```{r visualize-widening}
# Plot original and widened channels
par(mfrow = c(1, 2))
plot(channel, main = "Original Channel")
plot(channel_widened, main = "Widened Channel")
par(mfrow = c(1, 1))
```

## Summary

This vignette covered:

- Basic channel widening by width and volume
- Simple and complex side specifications using side functions
- Section-wise side proportions via `xchan_side` objects
- Erosion volume and width calculations
- Variable widening patterns

The next vignette will show you how to perform detailed channel calculations and analysis.
