---
title: "Channels with Profiles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Channels with Profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```


If your workflow requires profile cross sections, `xchan` allows a profile view to be encoded in addition to a planimetric encoding.

## Creating profiles by hand

```{r width-widening}
widths <- c(10, 12, 8, 15, 11, 9)
depths <- c(1, 1, 3, 2, 2, 3)
channel <- xt_as_channel(units::set_units(widths, "m"))

topo <- function(width, depth, ...) {
  d_left <- -(10:1)
  d_right <- 1:10 + width
  d_inner <- c(0, width / 3, width)
  e_left <- 0.5 + 1.5 * log(10:1 + 1)
  e_right <- rep(0, 10)
  e_inner <- c(0.5, -depth, 0)
  d <- c(d_left, d_inner, d_right)
  e <- c(e_left, e_inner, e_right)
  data.frame(
    distance = units::set_units(d, "m"),
    elevation = units::set_units(e, "m"),
    is_bank = d == 0 | d == width,
    ...
  )
}

df_list <- mapply(
  topo,
  width = widths,
  depth = depths,
  id = seq_along(widths),
  SIMPLIFY = FALSE
)

profile <- do.call(rbind, df_list)

channel <- xt_add_profile(
  channel,
  distance = distance,
  elevation = elevation,
  section = id,
  banks = is_bank,
  data = profile,
)
```

```{r}
plot(channel[[3]], extent = "full")
```


## Generating profiles from spatial data

```{r}
chan <- xt_generate_plan(demo_bankline, spacing = units::set_units(100, "m"))

# Plot
plot(demo_bankline, fill = "#7cb2fe")
plot(chan, add = TRUE)
```

```{r}
xt_axis(chan)
```




So far, this channel is only planimetric, with no profile view. If a profile view is important to you, you can get that from a DEM, piggybacking off the [**terra**]() package. The `xt_generate_profile()` adds a profile view to a pre-existing channel using a DEM, allowing you to sample the DEM by set spacing (`sample_freq`) or desired number of points (`sample_n`).

```{r}
demo_demm <- terra::unwrap(demo_dem)
chann <- xt_generate_profile(
  chan,
  dem = demo_demm,
  sample_freq = units::set_units(2, m),
  progress = TRUE
)
```


While `plot()`ing a channel only ever shows the planimetric view, the profiles can be viewed by selecting individual cross sections.

```{r}
section <- xt_xsection_at(chan, 20)
section
plot(section, view = "profile")
```


## Profile Widening

Channel widening now also applies to the profile cross sections.

```{r}
channel2 <- xt_widen(channel, dw = c(2, 1, 3, 0, 0, 0))
plot(channel2, col = "blue")
plot(channel, col = "red", add = TRUE)
```


With a profile, this has the effect of shifting the between-channel topography on either side of the thalweg further from the thalweg, spreading out the channel bottom at the thalweg. 

Trace a channel axis from the cross-sections (if you want to build an axis from
a bankline polygon alone, use the **centerline** package on CRAN).

```{r}
# Widen by a set distance
channel_widened <- xt_widen(channel, dw = 4)

plot(channel_widened, col = "red")
plot(channel, add = TRUE, axis = "arrows")
```

```{r}
plot(channel[[4]], extent = "full")
plot(channel_widened[[4]], extent = "full")
```


This operation is vectorized, so different amounts of erosion can be specified for each cross section.

```{r width-widening}
# Create a simple channel
channel <- xt_as_channel(c(10, 12, 8, 15, 11, 9))

# Widen by adding 2 meters to each cross-section
channel_widened <- xt_widen(channel, dw = c(5, 3, 7, 0, 0, 0))

xt_width(channel_widened)
```


If a profile view is available, you can also widen by volume -- provided there is enough volume to erode in the first place.

Sometimes it's useful to know how much volume of erosion would result from a given change in width, or vice versa, the width change resulting from a volume change. Note that because volume is in squared length units, not cubed, because it corresponds to a cross section.

```{r}
xt_erosion_width(channel, dv = units::set_units(10, "m2"))
xt_erosion_volume(channel, dw = units::set_units(100, "m"))
```


## Exporting

Channels having a profile encoding have more export options. Like for planimetric-only channels, exports go through the [**sf**]() package, making an `"sfc"` object. In addition to the planimetric encoding, you can also choose an export of the profiles alone, or construct a 3D encoding of the channel.

```{r}
xt_as_sfc(channel)
```