---
title: "Working with Complex Channel Geometries"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Complex Channel Geometries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```

# Working with Complex Channel Geometries

This vignette covers working with complex channel geometries including islands, bars, and irregular channel shapes.

## Creating Profile Cross-Sections with Islands

### Simple Channel Profile

```{r simple-profile}
# Create coordinates for a simple channel
coords <- matrix(c(
  -10, 15,  # left bank
  -5, 12,
  -2, 8,
  0, 5,     # thalweg
  2, 8,
  5, 12,
  10, 15    # right bank
), ncol = 2, byrow = TRUE)

# Create simple channel profile (water from -3 to 3)
profile_simple <- xt_new_profile(coords, bankpoints = c(-3, 3))
profile_simple
```

### Channel with Island

```{r island-profile}
# Create coordinates for channel with island
coords_island <- matrix(c(
  -10, 15,  # left bank
  -5, 12,
  -3, 8,    # left water edge
  -1, 6,    # island left edge
  1, 6,     # island right edge
  3, 8,     # right water edge
  5, 12,
  10, 15    # right bank
), ncol = 2, byrow = TRUE)

# Create profile with island (water from -3 to -1 and 1 to 3)
profile_island <- xt_new_profile(coords_island, bankpoints = c(-3, -1, 1, 3))
profile_island
```

### Multiple Islands

```{r multiple-islands}
# Create coordinates for channel with multiple islands
coords_multi <- matrix(c(
  -12, 15,  # left bank
  -8, 12,
  -6, 8,    # left water edge
  -4, 6,    # island 1 left
  -2, 6,    # island 1 right
  0, 8,     # water
  2, 6,     # island 2 left
  4, 6,     # island 2 right
  6, 8,     # right water edge
  8, 12,
  12, 15    # right bank
), ncol = 2, byrow = TRUE)

# Create profile with multiple islands
profile_multi <- xt_new_profile(coords_multi, bankpoints = c(-6, -4, -2, 0, 2, 4, 6))
profile_multi
```

## Active Width vs Total Width

### Understanding the Difference

```{r width-comparison}
# Create a channel with islands
channel_with_islands <- xt_as_channel(c(10, 12, 8, 15, 11, 9))

# Compare total width vs active width
total_widths <- xt_width(channel_with_islands)
# active_widths <- xt_width_active(channel_with_islands)

# The difference represents islands/land areas
# island_widths <- total_widths - active_widths

# data.frame(
#   total = total_widths,
#   active = active_widths,
#   islands = island_widths
# )
```

## Bank Analysis

### Spanning Banks

```{r bank-spanning}
# Analyze bank characteristics
# bank_analysis <- xt_span_banks(channel_with_islands)
# bank_analysis
```

## Complex Bankline Tracing

### Different Tracer Methods

```{r tracer-methods}
# Linear tracer (connects bank points directly)
linear_tracer <- tracer_linear()

# Spline tracer (smooth interpolation)
spline_tracer <- tracer_spline(degree = 3, smoothing = 0.5)

# DEM tracer (tracing contours from DEM)
# dem_tracer <- tracer_dem(my_dem, contour_interval = 1)
```

### Using Tracers (Operator Factories)

Tracers are **operator factories** that create functions for connecting bank points into continuous banklines. Similarly, extenders are **operator factories** that create functions for extending the topographic frame beyond the banks.

```{r using-tracers}
# Generate banklines using different tracers
# linear_banks <- xt_trace_banks(channel, tracer = linear_tracer)
# spline_banks <- xt_trace_banks(channel, tracer = spline_tracer)

# Or use text shortcuts
# linear_banks <- xt_trace_banks(channel, tracer = "linear")
# spline_banks <- xt_trace_banks(channel, tracer = "spline")
```

### Custom Tracers

```{r custom-tracer}
# Create a custom tracer for specific scenarios
custom_tracer <- function(channel) {
  # Custom bankline generation logic
  # This would return bankline polygons
  # For example, tracing along specific elevation contours
  NULL  # Placeholder
}

# Use custom tracer
# custom_banks <- xt_trace_banks(channel, tracer = custom_tracer)
```

## Working with Irregular Channel Shapes

### Asymmetric Channels

```{r asymmetric}
# Create asymmetric channel coordinates
coords_asym <- matrix(c(
  -8, 15,   # left bank
  -4, 12,
  -2, 8,
  0, 5,     # thalweg (off-center)
  3, 8,
  6, 12,
  12, 15    # right bank (wider)
), ncol = 2, byrow = TRUE)

# Create asymmetric profile
profile_asym <- xt_new_profile(coords_asym, bankpoints = c(-3, 4))
profile_asym
```

### Braided Channels

```{r braided}
# Create coordinates for braided channel
coords_braided <- matrix(c(
  -15, 15,  # left bank
  -10, 12,
  -8, 8,    # left channel
  -6, 6,    # bar 1 left
  -4, 6,    # bar 1 right
  -2, 8,    # center channel
  0, 6,     # bar 2 left
  2, 6,     # bar 2 right
  4, 8,     # right channel
  6, 12,
  10, 15    # right bank
), ncol = 2, byrow = TRUE)

# Create braided profile
profile_braided <- xt_new_profile(coords_braided, bankpoints = c(-8, -6, -4, -2, 0, 2, 4))
profile_braided
```

## Visualizing Complex Geometries

### Plotting Profiles with Islands

```{r plot-complex}
# Plot different profile types
par(mfrow = c(2, 2))

# Simple channel
plot(profile_simple, main = "Simple Channel")

# Channel with island
plot(profile_island, main = "Channel with Island")

# Multiple islands
plot(profile_multi, main = "Multiple Islands")

# Asymmetric channel
plot(profile_asym, main = "Asymmetric Channel")

par(mfrow = c(1, 1))
```

### 3D Visualization

```{r 3d-visualization}
# Create channel with complex profiles
# channel_complex <- xt_channel(.plan = plan_sfc, .profile = list(profile_island, profile_multi))

# Plot 3D view
# plot(channel_complex, view = "3d", exaggerate = 2)
```

## Advanced Island Analysis

### Island Characteristics

```{r island-analysis}
# Analyze island properties
# island_analysis <- function(profile) {
#   # Extract island information from profile
#   # Calculate island width, height, area, etc.
#   # This would be a custom analysis function
# }

# Apply to different profiles
# island_simple <- island_analysis(profile_island)
# island_multi <- island_analysis(profile_multi)
```

### Channel Complexity Metrics

```{r complexity-metrics}
# Calculate channel complexity metrics
complexity_metrics <- function(channel) {
  # Calculate various complexity measures
  # - Number of islands per cross-section
  # - Island area ratio
  # - Channel sinuosity
  # - Width variability
  # This would be a custom analysis function
}

# Apply to channel
# metrics <- complexity_metrics(channel_with_islands)
```

## Summary

This vignette covered:

- Creating profile cross-sections with islands using `xt_new_profile()`
- Understanding active width vs total width with `xt_width_active()`
- Bank analysis with `xt_span_banks()`
- Complex bankline tracing with different tracers
- Working with irregular channel shapes (asymmetric, braided)
- Visualizing complex geometries
- Advanced analysis of islands and channel complexity

These techniques allow you to work with realistic river channel geometries that include islands, bars, and other complex features commonly found in natural river systems.
