---
title: "Getting Started with xchan"
author: "Vincenzo Coia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with xchan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```

# Getting Started with xchan

The `xchan` package provides tools for creating and manipulating river channel cross-sections. This vignette will get you started with the basics.

## Creating Your First Channel Object

The core data structure in `xchan` is the **channel object**. A channel contains planimetric cross-sections (top-down view) and/or profile cross-sections (side view).

### Simple Channel from Widths

Let's get set up with a quick, simple channel. The simplest way to do this is by specifying a vector of widths corresponding to channel cross sections.

```{r simple-channel}
# Create a channel with 6 cross-sections of varying widths
widths <- c(10, 12, 8, 15, 11, 9)
channel <- xt_as_channel(widths)
channel
```

Notice that channels are simply data frames where each row is a cross section. So far there is only one column: the planimetric (overhead view) cross-sections created from the specified widths. Add other columns using your usual data frame workflow:

```{r df-channel}
# Add columns using ordinary data frame semantics
channel$roughness <- c(0.03, 0.035, 0.025, 0.04, 0.03, 0.025)
channel$station <- c(0, 100, 200, 300, 400, 500)
channel
```

Planimetric cross sections are sufficient for defining a channel, and are sufficient for this demo. You can also add a profile (side) view, and make more realistic channels; see vigette XYZ for more.

Here is what our channel looks like:

```{r}
plot(channel, view = "plan")
```

## Basic Channel Operations

Go ahead and widen the channel, and compare the new channel (blue) with the original (red).

```{r}
channel2 <- xt_widen(channel, width = c(2, 1, 3, 0, 0, 0)
plot(channel2, view = "plan", col = "blue")
plot(channel, view = "plan", col = "red", add = TRUE)
```

Once you have a channel object, you can perform various calculations. You can generate bankline polygons, calculate widths, and more. Remember, pretty much everything is customizable; see the various other vignettes for more.

Generate banklines and a centerline from the channel cross-sections (if you want to generate a centerline from a bankline polygon, use the `centerline` package found on CRAN).

```{r}
bankline <- xt_bankline(channel2)
centerline <- xt_centerline(channel2)
plot(bankline, view = "plan", col = "blue")
plot(channel2, view = "plan", col = "red", add = TRUE)
plot(centerline, view = "plan", add = TRUE)
```


Calculate cross-section widths.

```{r width-analysis}
# Get widths of all cross-sections
xt_width(channel)
```

Get the number of cross-sections.

```{r}
# Get number of cross-sections
xt_n_sections(channel)
```

Calculate downstream distances.

```{r distance}
# Calculate downstream distances
distances <- xt_distance_ds(channel)
distances
```

Convert to a spatial object understandable by the `sf` package, in case you want to do some manual operations not supported by `xchan`.

```{r}
spatial_object <- xt_as_sfc(channel)
# View
spatial_object
```

## Next Steps

Now that you understand the basics, you can:

1. **Create channels from different data sources** - See the "Channel Objects and Generation" vignette
2. **Modify channels through widening** - See the "Channel Widening and Erosion" vignette  
3. **Perform detailed calculations** - See the "Channel Calculations and Analysis" vignette
4. **Work with complex geometries** - See the "Working with Complex Channel Geometries" vignette

The channel object is the foundation of `xchan`. Once you're comfortable creating and manipulating channels, you can explore the more advanced features of the package.
