---
title: "Getting Started with xchan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with xchan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(xchan)
```

The `xchan` package provides tools for creating and manipulating river channel cross-sections. This vignette walks you through a typical workflow.


## Basic Channel Operations

Go ahead and widen the channel, and compare the new channel (blue) with the original (red).

```{r}
channel2 <- xt_widen(channel, dw = c(2, 1, 3, 0, 0, 0))
plot(channel2, col = "blue")
plot(channel, col = "red", add = TRUE)
```

Once you have a channel object, you can perform various calculations: widths,
downstream spacing, tracing an approximate axis through the cross-sections, and
more. Remember, pretty much everything is customizable; see the various other
vignettes for more.

Trace a channel axis from the cross-sections (if you want to build an axis from
a bankline polygon alone, use the **centerline** package on CRAN).

```{r}
axis_line <- xt_axis(channel2)
plot(channel2, col = "red")
plot(axis_line, add = TRUE, col = "blue", lwd = 2)
```


Calculate cross-section widths.

```{r width-analysis}
# Get widths of all cross-sections
xt_width(channel)
```

Get the number of cross-sections.

```{r}
# Get number of cross-sections
xt_n_sections(channel)
```

Calculate downstream distances.

```{r distance}
# Calculate downstream distances
distances <- xt_distance_downstream(channel)
distances
```

Convert to a spatial object understandable by the `sf` package, in case you want to do some manual operations not supported by `xchan`.

```{r}
spatial_object <- xt_as_sfc(channel)
# View
spatial_object
```

## Next Steps

Now that you understand the basics, you can:

1. **Create channels from different data sources** - See the "Channel Objects and Generation" vignette
2. **Modify channels through widening** - See the "Channel Widening and Erosion" vignette  
3. **Perform detailed calculations** - See the "Channel Calculations and Analysis" vignette
4. **Work with complex geometries** - See the "Working with Complex Channel Geometries" vignette

The channel object is the foundation of `xchan`. Once you're comfortable creating and manipulating channels, you can explore the more advanced features of the package.
