---
title: "Excluded prototypes"
subtitle: "Draft code and examples — not part of the installed package"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This document is **not** a package vignette: it is only stored in the source tree next to R prototypes under `excluded/`. Nothing here is exported or guaranteed to work with the current API.

# Dredging

Rectangular dredging and related entry points were moved out of the main package. See `dredgers.R`, `xt_dredge.R`, and `print.xchan_dredger.R` in this folder.

```{r dredging}
library(xchan)

# Requires profile cross-sections on the channel
# channel_dredged <- xt_dredge(channel, dredger = dredger_rectangle(depth = 2))

# rect_dredger <- dredger_rectangle(depth = 2, wse = elevation_bank())
# channel_rect <- xt_dredge(channel, dredger = rect_dredger)

# custom_dredger <- function(channel) { channel }
# channel_custom <- xt_dredge(channel, dredger = custom_dredger)
```

# Topographic frame extension

Extenders add synthetic topography beyond the banks for widening experiments. See `extenders.R`, `xt_extend_frame.R`, `xt_extend_frame_left.R`, `xt_extend_frame_right.R`, and `print.xchan_extender.R`.

```{r extenders}
# channel_extended <- xt_extend_frame(channel, extender = extender_flat(extent = 20))
# channel_sloped <- xt_extend_frame(channel, extender = extender_slope(extent = 30, slope = 0.02))
# channel_asymmetric <- xt_extend_frame_left(channel, extender = extender_slope(extent = 30, slope = 0.05))
# channel_asymmetric <- xt_extend_frame_right(channel_asymmetric, extender = extender_flat(extent = 20))
```

# Bankline tracers

Tracer factories and `xt_trace_banks()` connect bank points into continuous banklines. See `tracers.R`, `tracer_spline.R`, `tracer_concaveman.R`, and `xt_trace_banks.R`.

```{r tracers}
linear_tracer <- tracer_linear()
spline_tracer <- tracer_spline(degree = 3, smoothing = 0.5)

# linear_banks <- xt_trace_banks(channel, tracer = linear_tracer)
# spline_banks <- xt_trace_banks(channel, tracer = spline_tracer)
# linear_banks <- xt_trace_banks(channel, tracer = "linear")

custom_tracer <- function(channel) {
  NULL
}

# custom_banks <- xt_trace_banks(channel, tracer = custom_tracer)
```
